### 2023-02-03 products/manuscript/Manuscript.qmd
Submit Part 1

---
title: "MADA Final Project"
subtitle: "Trends in PrEP Use in the US, 2012-2022"
author: Nicole Luisi
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

# PART 1

**Overview**:

The purpose of this analysis is to measure the public health impact of pre-exposure prophylaxis (PrEP) in the US over the past 10 years. Previous research by [Sullivan et al. (2018)](https://doi.org/10.1016/j.annepidem.2018.06.009) evaluated trends in the prevalence of PrEP use from 2012-2017. Outcomes include period prevalence of PrEP use and PrEP-to-need ratio (PnR) by year. This analysis will extend the previous work to include data through the year 2022; subgroups will include region, state, age group, gender, and race/ethnicity *(race/ethnicity PrEP data were not available at the time of the previous analysis and were not included)*. [@SULLIVAN2018833]

**Data Source**:

*Note: New diagnoses data are currently available through 2020; PrEP data through 2021. If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*

+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020] 
+ PrEP and PnR: IQVIA (Quintiles IMS Holdings)
+ Annual Population Estimates: National Historical Geographic Information System (NHGIS), American Community Surveys (ACS), US Census Bureau

**Analysis**:

+ N/% of annual PrEP users by age, gender, race/ethnicity, and region
+ Annual prevalence of PrEP use per 100,000 population by age, gender, race/ethnicity, region, and state
+ Estimated annual percent change (EAPC) for prevalence
  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
+ Prep-to-need ratio by age, gender, race/ethnicity, and region
+ Other modeling TBD based on course modules

{{< pagebreak >}}

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
#library(here)
#library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in []

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by sex."
#| echo: FALSE
#knitr::include_graphics(here("results","height_weight_stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
#resulttable2 = readRDS(here("results","resulttable2.rds"))
#knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [] discusses types of analyses.

These papers [] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2023-02-24 code/processing_code/processingfile_v2.qmd
Submit Part 2

---
title: "An example cleaning script with code pulled in"
author: "Andreas Handel"
date: "2023-01-03"
output: html_document
---

# Processing script with code pulled in

This is essentially the same as the other Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script using the code chunk labels.

# Setup

This needs to run to load the R script so we can include the code chunks below.

Note that you can pull in more than one R script if you want.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages. Instead of having R commands here, this just pulls in the code from the R script we loaded above and labelled.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r, loaddata}
```

# Combine data

```{r, combinedata}
```

# Explore data

Several ways of looking at the data

```{r, exploredata}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

Looks like we have the following data:

* height (seems like it's in centimeters) 
* weight (seems to be in kilogram)
* Sex

# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number. 
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r, cleandata1}
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r, cleandata2}
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

Note that removing anyone who had "faulty" or missing data is one approach. It's often not the best. Based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

```{r, cleandata3}
```

Now checking the `Sex` variable. Sex should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.

```{r, cleandata4}
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Sex, I'm also using droplevels() to get rid of it.

```{r, cleandata5}
```

# Save data

All done, data is clean now. Time to save.
Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r, savedata}
```

### 2023-02-24 products/manuscript/Manuscript.qmd
Submit Part 2

@@
 title: "MADA Final Project"
-subtitle: "Trends in PrEP Use in the US, 2012-2022"
+subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
 author: Nicole Luisi
@@
 
-# PART 1
-
-**Overview**:
-
-The purpose of this analysis is to measure the public health impact of pre-exposure prophylaxis (PrEP) in the US over the past 10 years.
-Previous research by [Sullivan et al.
-(2018)](https://doi.org/10.1016/j.annepidem.2018.06.009) evaluated trends in the prevalence of PrEP use from 2012-2017.
-Outcomes include period prevalence of PrEP use and PrEP-to-need ratio (PnR) by year.
-This analysis will extend the previous work to include data through the year 2022; subgroups will include region, state, age group, gender, and race/ethnicity *(race/ethnicity PrEP data were not available at the time of the previous analysis and were not included)*.
-[@SULLIVAN2018833]
-
-**Data Source**:
-
-*Note: New diagnoses data are currently available through 2020; PrEP data through 2021.
-If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
-
-+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020] 
-+ PrEP and PnR: IQVIA (Quintiles IMS Holdings)
-+ Annual Population Estimates: National Historical Geographic Information System (NHGIS), American Community Surveys (ACS), US Census Bureau
-
-**Analysis**:
-
-+ N/% of annual PrEP users by age, gender, race/ethnicity, and region
-+ Annual prevalence of PrEP use per 100,000 population by age, gender, race/ethnicity, region, and state
-+ Estimated annual percent change (EAPC) for prevalence
-  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
-+ Prep-to-need ratio by age, gender, race/ethnicity, and region
-+ Other modeling TBD based on course modules
-
-{{< pagebreak >}}
-
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
+Note: There seem to be  lot of issues getting tables to render well in Word doc format, need to come back to this.
+I tried 5 or so different methods and they all look terrible when in a doc vs html.
+Also having issues with resolving crossrefs.
 
 ```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-#library(here)
-#library(knitr)
+# load R packages
+library(here) 
+library(ggplot2)
+library(skimr) 
+library(plotly)
+library(gt)
+library(knitr)
 ```
@@
 # Summary/Abstract
-_Write a summary of your project._
 
-{{< pagebreak >}}
+The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years.
+Previous research by [Chapin-Bardales et al.
+(2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
+This analysis will focus on new HIV diagnosesbetween  2008-2022; subgroups will include region, age group, gender, and race/ethnicity.
+[@Chapin2017]
 
@@
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
+**Data Source**:
+
+*Note: New diagnoses data are currently available through 2020.
+If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
+
++ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]
+
+**Analysis**:
+
++ Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
++ Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
++ Estimated annual percent change (EAPC) for incidence
+  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
++ Other modeling TBD based on course modules
 
@@
 _State the research questions you plan to answer with this analysis._
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in []
 
@@
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in the `data/raw_data/AIDSVu` subfolder.
+The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+The `processingcode.R` file includes the following steps:
++ Read in xlsx data files and combine all years for each level
++ Subset variables of interest and rename variables so all levels are consistent 
++ View/explore data
++ Calculate new rate ratio variables for subgroups 
++ Prepare combined national and regional file with all variables
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+@tbl-natsummarytable shows a full summary of the National data.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: tbl-natsummarytable
+#| tbl-cap: "National data summary table."
 #| echo: FALSE
-#resulttable=readRDS("../../results/summarytable.rds")
-#knitr::kable(resulttable)
+alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
+national<-alldata %>% filter(Geo %in% c("United States"))
+national %>% gt::gt() %>% 
+  gt::tab_header(title = "National Data") %>%
+  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
+                             gt::cell_text(weight = "bold")),
+                locations = gt::cells_body(columns = Year))
+```
+@fig-natplots shows exploratory plots for the national rates by subgroup.
+
+```{r}
+#| label: fig-natplots
+#| fig-cap: "National rates by subgroup"
+#| echo: FALSE
+national<-alldata %>% filter(Geo %in% c("United States"))
+natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
+natfig <- natfig %>% add_trace(y = ~Male_Rate, name = "Male", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
+natfig <- natfig %>% add_trace(y = ~Female_Rate, name = "Female", mode = 'lines', line = list(width = 3, dash = "solid", color = "green"))
+natfig %>% layout(title="National New Diagnosis Rate by Sex", legend=list(title=list(text='Sex')), 
+                  xaxis = list(title = 'Year'),
+                  yaxis = list(title = 'Rate per 100,000'))
+
+natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
+natfig <- natfig %>% add_trace(y = ~Black_Rate, name = "Black", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
+natfig <- natfig %>% add_trace(y = ~White_Rate, name = "White", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
+natfig <- natfig %>% add_trace(y = ~Hispanic_Rate, name = "Hispanic", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
+natfig <- natfig %>% add_trace(y = ~Asian_Rate, name = "Asian", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
+natfig <- natfig %>% add_trace(y = ~AIAN_Rate, name = "AIAN", mode = 'lines', line = list(width = 3, dash = "solid", color = "brown1"))
+natfig <- natfig %>% add_trace(y = ~MultRace_Rate, name = "Multiracial", mode = 'lines', line = list(width = 3, dash = "solid", color = "purple"))
+natfig <- natfig %>% add_trace(y = ~NHPI_Rate, name = "NHPI", mode = 'lines', line = list(width = 3, dash = "solid", color = "magenta"))
+natfig %>% layout(title="National New Diagnosis Rate by Race/Ethnicity", legend=list(title=list(text='Race/Ethnicity')), 
+                  xaxis = list(title = 'Year'),
+                  yaxis = list(title = 'Rate per 100,000'))
+
+natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
+natfig <- natfig %>% add_trace(y = ~Age13t24_Rate, name = "Age 13-24", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
+natfig <- natfig %>% add_trace(y = ~Age25t34_Rate, name = "Age 25-34", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
+natfig <- natfig %>% add_trace(y = ~Age35t44_Rate, name = "Age 35-44e", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
+natfig <- natfig %>% add_trace(y = ~Age45t54_Rate, name = "Age 45-54", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
+natfig <- natfig %>% add_trace(y = ~Age55p_Rate, name = "Age 55+", mode = 'lines', line = list(width = 3, dash = "solid", color = "orange"))
+natfig %>% layout(title="National New Diagnosis Rate by Age", legend=list(title=list(text='Age')), 
+                  xaxis = list(title = 'Year'),
+                  yaxis = list(title = 'Rate per 100,000'))
+```
+
+@tbl-natratios shows a summary of selected rate ratios for race/ethnicity and sex.
+
+```{r}
+#| label: tbl-natratios
+#| tbl-cap: "National rate ratios by select subgroups"
+#| echo: FALSE
+national<-alldata %>% filter(Geo %in% c("United States"))
+national_sub <- national %>% dplyr::select("Year", "Black_Rate", "White_Rate", "Hispanic_Rate", "black_white_rateratio", "hispanic_white_rateratio", "Male_Rate", "Female_Rate", "female_male_rateratio")
+national_sub %>% gt::gt() %>% 
+  gt::tab_header(title = "Rate Ratios") %>%
+  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
+                             gt::cell_text(weight = "bold")),
+                locations = gt::cells_body(columns = Year))%>%
+  gt::tab_style(style = list(gt::cell_fill(color = "#F1EB9C"),
+                             gt::cell_text(weight = "bold")), 
+                locations = gt::cells_body(columns = c(black_white_rateratio,hispanic_white_rateratio,female_male_rateratio)))
+```
+
+@tbl-regsummarytable shows a full summary of the Regional data.
+
+```{r}
+#| label: tbl-regsummarytable
+#| tbl-cap: "Regional data summary table."
+#| echo: FALSE
+alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
+regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
+regional %>% gt::gt() %>% 
+  gt::tab_header(title = "Regional Data") %>%
+  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
+                             gt::cell_text(weight = "bold")),
+                locations = gt::cells_body(columns = Year))
+```
+
+@fig-regplots shows an exploratory plot for regional overall new diagnosis rates with female-male rate ratios.
+
+```{r}
+#| label: fig-regplots
+#| tbl-cap: "Regional plots"
+#| echo: FALSE
+alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
+regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
+ggplot(regional, aes(x=Year, y=log(Overall_Rate), size=log(female_male_rateratio), color=Geo)) +
+    geom_point(alpha=0.7) + scale_x_continuous(breaks=c(2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020)) +
+  ggtitle("Regional New Diagnosis Rates with Female-Male Rate Ratios") +
+  theme(plot.title = element_text(hjust = 0.5)) +
+  xlab("Year") + ylab("Log of Overall NewDx Rate")
 ```

### 2023-02-24 code/processing_code/processingcode.R

@@ -16,124 +16,112 @@ library(dplyr) #for data processing/cleaning
 library(tidyr) #for data processing/cleaning
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
+library(plotly)
+library(gt)
 
 ## ---- loaddata --------
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","exampledata.xlsx")
+nat_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows  
+reg_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "Region", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
+#st_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "State", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
 
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
+#nat_pnr_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "PNR")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
+#nat_prep_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "PrEP")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
+#reg_pnr_files <- list.files(path = (here("data","raw_data", "AIDSVu", "Region", "PNR")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
[…]

### 2023-03-17 code/analysis_code/analysis.qmd
Submit Part 3

---
title: "Analysis script"
author: "Nicole Luisi"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r, echo=FALSE, message=FALSE}
# load R packages
library(here) 
library(readxl)
library(ggplot2)
library(skimr) 
library(plotly)
library(gt)
library(knitr)
library(dplyr)
library(kableExtra)
library(gridExtra)
library(ggpubr)
library(janitor)
```

# Load clean data

```{r}
alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
```

```{r, echo=FALSE, message=FALSE}
# Plot of US rates by sex, age, race/ethnicity 
national<-alldata %>% filter(Geo %in% c("United States"))
natfig_s <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig_s <- natfig_s %>% add_trace(y = ~Male_Rate, name = "Male", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig_s <- natfig_s %>% add_trace(y = ~Female_Rate, name = "Female", mode = 'lines', line = list(width = 3, dash = "solid", color = "green"))
natfig_s <- natfig_s %>% layout(title="National New Diagnosis Rate by Sex", legend=list(title=list(text='Sex')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))

natfig_r <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig_r <- natfig_r %>% add_trace(y = ~Black_Rate, name = "Black", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig_r <- natfig_r %>% add_trace(y = ~White_Rate, name = "White", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
natfig_r <- natfig_r %>% add_trace(y = ~Hispanic_Rate, name = "Hispanic", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
natfig_r <- natfig_r %>% add_trace(y = ~Asian_Rate, name = "Asian", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
natfig_r <- natfig_r %>% add_trace(y = ~AIAN_Rate, name = "AIAN", mode = 'lines', line = list(width = 3, dash = "solid", color = "brown1"))
natfig_r <- natfig_r %>% add_trace(y = ~MultRace_Rate, name = "Multiracial", mode = 'lines', line = list(width = 3, dash = "solid", color = "purple"))
natfig_r <- natfig_r %>% add_trace(y = ~NHPI_Rate, name = "NHPI", mode = 'lines', line = list(width = 3, dash = "solid", color = "magenta"))
natfig_r <- natfig_r %>% layout(title="National New Diagnosis Rate by Race/Ethnicity", legend=list(title=list(text='Race/Ethnicity')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))

natfig_a <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig_a <- natfig_a %>% add_trace(y = ~Age13t24_Rate, name = "Age 13-24", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig_a <- natfig_a %>% add_trace(y = ~Age25t34_Rate, name = "Age 25-34", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
natfig_a <- natfig_a %>% add_trace(y = ~Age35t44_Rate, name = "Age 35-44e", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
natfig_a <- natfig_a %>% add_trace(y = ~Age45t54_Rate, name = "Age 45-54", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
natfig_a <- natfig_a %>% add_trace(y = ~Age55p_Rate, name = "Age 55+", mode = 'lines', line = list(width = 3, dash = "solid", color = "orange"))
natfig_a <- natfig_a %>% layout(title="National New Diagnosis Rate by Age", legend=list(title=list(text='Age')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))

natfig_s
plotly::export(p=natfig_s, file = here("results", "US_rates_bysex.png"))
natfig_a
plotly::export(p=natfig_a, file = here("results", "US_rates_byage.png"))
natfig_r
plotly::export(p=natfig_r, file = here("results", "US_rates_byrace.png"))
```

# Load data from external Joinpoint program

```{r, echo=FALSE, message=FALSE}
# Read in data from external Joinpoint program

# Function to pull names of sheets in excel files (just applying to one since all are the same)

multiplesheets <- function(fname) {
  sheets <- readxl::excel_sheets(fname)
  tibble <- lapply(sheets, function(x) readxl::read_excel(fname, sheet = x))
  data_frame <- lapply(tibble, as.data.frame)
  names(data_frame) <- sheets
  print(data_frame)
}
path <- here("data", "processed_data", "djp_overall.xlsx")
multiplesheets(path)

# Overall
jpoutput_overall_aapc <- read_excel(here("data", "processed_data", "djp_overall.xlsx"), sheet="AAPCs")
names(jpoutput_overall_aapc)<- c("geo", "joinpoint_model", "AAPC_Index", "start", "end", "AAPC", "AAPC_lCI", "AAPC_uCI","significant", "test_statstic", "p-value")
jpoutput_overall_apc <- read_excel(here("data", "processed_data", "djp_overall.xlsx"), sheet="APCs")
names(jpoutput_overall_apc)<-c("Geo","Model","Segment","Segment Start","Segment End","APC","APC 95% LCL","APC 95% UCL","APC Significant","Test Statistic","P-Value")

# Race
jpoutput_race_aapc <- read_excel(here("data", "processed_data", "djp_race.xlsx"), sheet="AAPCs")
names(jpoutput_race_aapc)<- c("type","geo", "joinpoint_model", "AAPC_Index", "start", "end", "AAPC", "AAPC_lCI", "AAPC_uCI","significant", "test_statstic", "p-value")
jpoutput_race_data <- read_excel(here("data", "processed_data", "djp_race.xlsx"), sheet="Data")
names(jpoutput_race_data)<- c("type","geo", "year", "rate", "jp", "flag")
jpoutput_race_apc <- read_excel(here("data", "processed_data", "djp_race.xlsx"), sheet="APCs")
names(jpoutput_race_apc)<-c("Type","Geo","Model","Segment","Segment Start","Segment End","APC","APC 95% LCL","APC 95% UCL","APC Significant","Test Statistic","P-Value")

# Race Ratios
jpoutput_raceratios_aapc <- read_excel(here("data", "processed_data", "djp_raceratios.xlsx"), sheet="AAPCs")
names(jpoutput_raceratios_aapc)<- c("type","geo", "joinpoint_model", "AAPC_Index", "start", "end", "AAPC", "AAPC_lCI", "AAPC_uCI","significant", "test_statstic", "p-value")
jpoutput_raceratio_data <- read_excel(here("data", "processed_data", "djp_raceratios.xlsx"), sheet="Data")
names(jpoutput_raceratio_data)<- c("type","geo", "year", "rate", "jp", "flag")
jpoutput_raceratio_apc <- read_excel(here("data", "processed_data", "djp_raceratios.xlsx"), sheet="APCs")
names(jpoutput_raceratio_apc)<-c("Type","Geo","Model","Segment","Segment Start","Segment End","APC","APC 95% LCL","APC 95% UCL","APC Significant","Test Statistic","P-Value")

# Age
jpoutput_age_aapc <- read_excel(here("data", "processed_data", "djp_age.xlsx"), sheet="AAPCs")
names(jpoutput_age_aapc)<- c("type","geo", "joinpoint_model", "AAPC_Index", "start", "end", "AAPC", "AAPC_lCI", "AAPC_uCI","significant", "test_statstic", "p-value")
jpoutput_age_apc <- read_excel(here("data", "processed_data", "djp_age.xlsx"), sheet="APCs")
names(jpoutput_age_apc)<-c("Type","Geo","Model","Segment","Segment Start","Segment End","APC","APC 95% LCL","APC 95% UCL","APC Significant","Test Statistic","P-Value")

# Sex
jpoutput_sex_aapc <- read_excel(here("data", "processed_data", "djp_sex.xlsx"), sheet="AAPCs")
names(jpoutput_sex_aapc)<- c("type","geo", "joinpoint_model", "AAPC_Index", "start", "end", "AAPC", "AAPC_lCI", "AAPC_uCI","significant", "test_statstic", "p-value")
jpoutput_sex_apc <- read_excel(here("data", "processed_data", "djp_sex.xlsx"), sheet="APCs")
names(jpoutput_sex_apc)<-c("Type","Geo","Model","Segment","Segment Start","Segment End","APC","APC 95% LCL","APC 95% UCL","APC Significant","Test Statistic","P-Value")
```

## Table of AAPCs

```{r, echo=FALSE, message=FALSE}
jpoutput_overall_aapc$type<-"Overall"
jpoutput_overall_aapc <- jpoutput_overall_aapc[, c(12, 1:11)]
all_aapc<-rbind(jpoutput_overall_aapc,jpoutput_sex_aapc,jpoutput_age_aapc,jpoutput_race_aapc,jpoutput_raceratios_aapc) %>% select("type","geo", "AAPC", "AAPC_lCI", "AAPC_uCI", "p-value")

kable(all_aapc,caption = "All AAPC Results", 
      col.names=c("Stratification", "Geographic Area", "AAPC", "AAPC_lCI", "AAPC_uCI", "p-value"),      align="rc") %>% 
  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
  pack_rows("Overall", 1, 5) %>%
  pack_rows("Female", 6, 10) %>%
  pack_rows("Male", 11, 15) %>%
  pack_rows("Age 13-24", 16, 20) %>%
  pack_rows("Age 25-34", 21, 25) %>%
  pack_rows("Age 35-44", 26, 30)  %>%
  pack_rows("Age 45-54", 31, 35)  %>%
  pack_rows("Age 55+", 36, 40)  %>%
  pack_rows("American Indian/Alaska Native", 41, 44) %>%
  pack_rows("Asian", 45, 49) %>%
  pack_rows("Black", 50, 54) %>%
  pack_rows("Hispanic", 55, 59) %>%
  pack_rows("Multiracial", 60, 64) %>%
  pack_rows("Native Hawaiian/Pacific Islander", 65, 67) %>%
  pack_rows("White", 68, 72) %>%
  pack_rows("AIAN-White", 73, 76) %>%
  pack_rows("Asian-White", 77, 81) %>%
  pack_rows("Black-White", 82, 86) %>%
  pack_rows("Hispanic-White", 87, 91) %>%
  pack_rows("Multi-White", 92, 96) %>%
  pack_rows("NHPI-White", 97, 99)

table_file6 = here("results", "aapc_all_results.rds")
saveRDS(all_aapc, file = table_file6)
```

## Plot of US Race Rate Ratios (White=ref)

```{r, echo=FALSE, message=FALSE}
US_jpoutput_raceratios_aapc<-jpoutput_raceratios_aapc %>% filter(geo %in% c("United States"))
fig_US_raceratio_AAPC<-ggplot(US_jpoutput_raceratios_aapc, aes(type, AAPC)) +   
      geom_point() + theme_minimal() + ggtitle("US") + theme(plot.title = element_text(hjust = 0.5)) +
      geom_errorbar(aes(ymin = AAPC_lCI, ymax = AAPC_uCI)) +
      scale_x_discrete(labels=c('AIAN', 'Asian', 'Black', 'Hispanic', 'Multi', 'NHPI')) +
      xlab("Race Ratio (vs White)") + ylab("Avg Annual % Change")

South_jpoutput_raceratios_aapc<-jpoutput_raceratios_aapc %>% filter(geo %in% c("South"))
fig_South_raceratio_AAPC<-ggplot(South_jpoutput_raceratios_aapc, aes(type, AAPC)) +   
      geom_point() + theme_minimal() + ggtitle("South") + theme(plot.title = element_text(hjust = 0.5)) +
      geom_errorbar(aes(ymin = AAPC_lCI, ymax = AAPC_uCI)) +
      scale_x_discrete(labels=c('AIAN', 'Asian', 'Black', 'Hispanic', 'Multi', 'NHPI')) +
      xlab("Race Ratio (vs White)") + ylab("Avg Annual % Change")

Northeast_jpoutput_raceratios_aapc<-jpoutput_raceratios_aapc %>% filter(geo %in% c("Northeast"))
fig_Northeast_raceratio_AAPC<-ggplot(Northeast_jpoutput_raceratios_aapc, aes(type, AAPC)) +        
      geom_point() + theme_minimal() + ggtitle("Northeast") + theme(plot.title = element_text(hjust = 0.5)) +
      geom_errorbar(aes(ymin = AAPC_lCI, ymax = AAPC_uCI)) +
      scale_x_discrete(labels=c('Asian', 'Black', 'Hispanic', 'Multi')) +
      xlab("Race Ratio (vs White)") + ylab("Avg Annual % Change")

West_jpoutput_raceratios_aapc<-jpoutput_raceratios_aapc %>% filter(geo %in% c("West"))
fig_West_raceratio_AAPC<-ggplot(West_jpoutput_raceratios_aapc, aes(type, AAPC)) + 
      geom_point() + theme_minimal() + ggtitle("West") + theme(plot.title = element_text(hjust = 0.5)) +
      geom_errorbar(aes(ymin = AAPC_lCI, ymax = AAPC_uCI)) +
      scale_x_discrete(labels=c('AIAN', 'Asian', 'Black', 'Hispanic', 'Multi', 'NHPI')) +
      xlab("Race Ratio (vs White)") + ylab("Avg Annual % Change")

Midwest_jpoutput_raceratios_aapc<-jpoutput_raceratios_aapc %>% filter(geo %in% c("Midwest"))
fig_Midwest_raceratio_AAPC<-ggplot(Midwest_jpoutput_raceratios_aapc, aes(type, AAPC)) + 
      geom_point() + theme_minimal() + ggtitle("Midwest") + theme(plot.title = element_text(hjust = 0.5)) +
      geom_errorbar(aes(ymin = AAPC_lCI, ymax = AAPC_uCI)) +
      scale_x_discrete(labels=c('AIAN', 'Asian', 'Black', 'Hispanic', 'Multi')) +
      xlab("Race Ratio (vs White)") + ylab("Avg Annual % Change")

rr_reg<-grid.arrange(arrangeGrob(fig_Northeast_raceratio_AAPC, fig_South_raceratio_AAPC, fig_Midwest_raceratio_AAPC, fig_West_raceratio_AAPC, nrow = 2))

rr_reg
figure_file = here("results","raceratio_byregion.png")
ggsave(filename = figure_file, plot=rr_reg) 
```

## Plot of US Rates by Race

```{r, echo=FALSE, message=FALSE}
US_jpoutput_race<-jpoutput_race_data %>% filter(geo %in% c("United States"))
US_race_rate_fig <- ggplot(US_jpoutput_race, aes(x=year, y=log(rate), size=log(rate), color=type)) + theme_light() +
    geom_point(alpha=0.5) +
    scale_y_continuous("Log Rate") +
    scale_x_continuous("Year", breaks=c(2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020)) + ggtitle("US Rates by Race") + 
    theme(plot.title = element_text(hjust = 0.5)) + guides(color=guide_legend(title="Race/Ethnicity"), size=guide_legend(title="Log Rate"))

US_race_rate_fig
figure_file = here("results","racerates_allUS.png")
ggsave(filename = figure_file, plot=US_race_rate_fig) 
```

## Plot of Regional Rates by Race

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Northeast
Northeast_jpoutput_race<-jpoutput_race_data %>% filter(geo %in% c("Northeast"))
Northeast_race_rate_fig <- ggplot(Northeast_jpoutput_race, aes(x=year, y=log(rate), size=log(rate), color=type)) + theme_light() +
    geom_point(alpha=0.5) +
    scale_y_continuous("Log Rate") +
    scale_x_continuous("Year", breaks=c(2008,2010,2012,2014,2016,2018,2020)) + ggtitle("Northeast") + 
    theme(plot.title = element_text(hjust = 0.5)) + guides(color=guide_legend(title="Race/Ethnicity"), size=guide_legend(title="Log Rate"))

# South
South_jpoutput_race<-jpoutput_race_data %>% filter(geo %in% c("South"))
South_race_rate_fig <- ggplot(South_jpoutput_race, aes(x=year, y=log(rate), size=log(rate), color=type)) + theme_light() +
    geom_point(alpha=0.5) +
    scale_y_continuous("Log Rate") +
    scale_x_continuous("Year", breaks=c(2008,2010,2012,2014,2016,2018,2020)) + ggtitle("South") + 
    theme(plot.title = element_text(hjust = 0.5)) + guides(color=guide_legend(title="Race/Ethnicity"), size=guide_legend(title="Log Rate"))

# Midwest
Midwest_jpoutput_race<-jpoutput_race_data %>% filter(geo %in% c("Midwest"))
Midwest_race_rate_fig <- ggplot(Midwest_jpoutput_race, aes(x=year, y=log(rate), size=log(rate), color=type)) + theme_light() +
    geom_point(alpha=0.5) +
    scale_y_continuous("Log Rate") +
    scale_x_continuous("Year", breaks=c(2008,2010,2012,2014,2016,2018,2020)) + ggtitle("Midwest") + 
    theme(plot.title = element_text(hjust = 0.5)) + guides(color=guide_legend(title="Race/Ethnicity"), size=guide_legend(title="Log Rate"))

# West
West_jpoutput_race<-jpoutput_race_data %>% filter(geo %in% c("West"))
West_race_rate_fig <- ggplot(West_jpoutput_race, aes(x=year, y=log(rate), size=log(rate), color=type)) + theme_light() +
    geom_point(alpha=0.5) +
    scale_y_continuous("Log Rate") +
    scale_x_continuous("Year", breaks=c(2008,2010,2012,2014,2016,2018,2020)) + ggtitle("West") + 
    theme(plot.title = element_text(hjust = 0.5)) + guides(color=guide_legend(title="Race/Ethnicity"), size=guide_legend(title="Log Rate"))

r_reg<-ggarrange(Northeast_race_rate_fig, South_race_rate_fig, Midwest_race_rate_fig, West_race_rate_fig, ncol=2, nrow = 2, common.legend = TRUE, legend="right")

r_reg
figure_file = here("results","racerates_region.png")
ggsave(filename = figure_file, plot=r_reg) 
```

## Joinpoint Plots for Race

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Black
US_jpoutput_black<-jpoutput_race_data %>% filter(geo %in% c("United States") & type %in% c("Black"))
black_jp_plot <- ggplot(US_jpoutput_black, aes(x=year, y=rate)) + geom_point() + 
  theme_light() + geom_line(aes(year, as.numeric(jp)), color="blue") +
  scale_y_continuous("Rate", breaks = c(0, 20, 40, 60, 80), limits=c(0, 80)) +
  scale_x_continuous("Year", breaks=c(2008,2011,2014,2017,2020)) + ggtitle("All US: Black") + theme(plot.title = element_text(hjust = 0.5))

# White
US_jpoutput_white<-jpoutput_race_data %>% filter(geo %in% c("United States") & type %in% c("White"))
white_jp_plot <- ggplot(US_jpoutput_white, aes(x=year, y=rate)) + geom_point() + 
  theme_light() + geom_line(aes(year, as.numeric(jp)), color="blue") +
  scale_y_continuous("Rate", breaks = c(0, 20, 40, 60, 80), limits=c(0, 80)) +
  scale_x_continuous("Year", breaks=c(2008,2011,2014,2017,2020)) + ggtitle("All US: White") + theme(plot.title = element_text(hjust = 0.5))

# Hispanic
US_jpoutput_hisp<-jpoutput_race_data %>% filter(geo %in% c("United States") & type %in% c("Hispanic"))
hisp_jp_plot <- ggplot(US_jpoutput_hisp, aes(x=year, y=rate)) + geom_point() + 
  theme_light() + geom_line(aes(year, as.numeric(jp)), color="blue") +
  scale_y_continuous("Rate", breaks = c(0, 20, 40, 60, 80), limits=c(0, 80)) +
  scale_x_continuous("Year", breaks=c(2008,2011,2014,2017,2020)) + ggtitle("All US: Hispanic") + theme(plot.title = element_text(hjust = 0.5))

jp_all_bwh<-grid.arrange(arrangeGrob(black_jp_plot, hisp_jp_plot, white_jp_plot, nrow = 1))

jp_all_bwh
figure_file = here("results","raceBWH_joinpoints_allUS.png")
ggsave(filename = figure_file, plot=jp_all_bwh) 
```

## Supplemental Table of APCs

```{r, echo=FALSE, message=FALSE}
jpoutput_overall_apc$Type<-"Overall" 
jpoutput_overall_apc <- jpoutput_overall_apc[, c(12, 1:11)]
all_apc<-rbind(jpoutput_overall_apc,jpoutput_sex_apc,jpoutput_age_apc,jpoutput_race_apc,jpoutput_raceratio_apc)

kable(all_apc,caption = "All APC Results", align="rc") %>% 
  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
  pack_rows("Overall", 1, 15) %>%
  pack_rows("Female", 16, 24) %>%
  pack_rows("Male", 25, 32) %>%
  pack_rows("Age 13-24", 33, 41) %>%
  pack_rows("Age 25-34", 42, 54) %>%
  pack_rows("Age 35-44", 55, 63)  %>%
  pack_rows("Age 45-54", 64, 70)  %>%
  pack_rows("Age 55+", 71, 76)  %>%
  pack_rows("American Indian/Alaska Native", 77, 80) %>%
  pack_rows("Asian", 81, 90) %>%
  pack_rows("Black", 91, 102) %>%
  pack_rows("Hispanic", 103, 113) %>%
  pack_rows("Multiracial", 114, 120) %>%
  pack_rows("Native Hawaiian/Pacific Islander", 121, 123) %>%
  pack_rows("White", 124, 130) %>%
  pack_rows("AIAN-White", 131, 134) %>%
  pack_rows("Asian-White", 135, 142) %>%
  pack_rows("Black-White", 143, 147) %>%
  pack_rows("Hispanic-White", 148, 156) %>%
  pack_rows("Multi-White", 157, 165) %>%
  pack_rows("NHPI-White", 166, 168)

table_file7 = here("results", "apc_all_results.rds")
saveRDS(all_apc, file = table_file7)
```

## Rate difference comparisons by race (Black and Hispanic vs White)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ratediff_us_compare_bwh<-alldata %>% filter(Geo %in% c("United States")) %>%  dplyr::select(Year,Black_Rate,White_Rate,Hispanic_Rate, Asian_Rate, AIAN_Rate, MultRace_Rate,NHPI_Rate)
ratediff_us_compare_bwh$DiffWB<-ratediff_us_compare_bwh$White_Rate - ratediff_us_compare_bwh$Black_Rate
ratediff_us_compare_bwh$DiffWH<-ratediff_us_compare_bwh$White_Rate - ratediff_us_compare_bwh$Hispanic_Rate
nat_diff_bwh <- plot_ly(ratediff_us_compare_bwh, x = ~Year, y = ~DiffWB, name = "White-Black Rate Difference", type="scatter", mode = "lines", line = list(width = 3, dash = "solid", color = "blue")) 
nat_diff_bwh <- nat_diff_bwh %>% add_trace(y = ~DiffWH, name = "White-Hispanic Rate Difference", mode = 'lines', line = list(width = 3, dash = "solid", color = "green"))
nat_diff_bwh <- nat_diff_bwh %>% layout(title="New Diagnoses Rate Differences by Year, all US", legend=list(title=list(text=' ')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate Difference'))

nat_diff_bwh
plotly::export(p=nat_diff_bwh, file = here("results", "nat_diff_bwh.png"))
```

## Supplemental Table of Rates, National

```{r, echo=FALSE, message=FALSE}
national<-alldata %>% filter(Geo %in% c("United States")) %>% dplyr::select(Year,Overall_Rate,Male_Rate,Female_Rate,Black_Rate,White_Rate,Hispanic_Rate,Asian_Rate,AIAN_Rate,MultRace_Rate,NHPI_Rate,Age13t24_Rate,Age25t34_Rate,Age35t44_Rate,Age45t54_Rate,Age55p_Rate)
national %>% gt::gt() %>% 
  gt::tab_header(title = "National Rate Data") %>%
  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
                             gt::cell_text(weight = "bold")),
                locations = gt::cells_body(columns = Year))

table_file8 = here("results", "all_us_rates.rds")
saveRDS(national, file = table_file8)
```

### 2023-03-17 code/analysis_code/exploratory_analysis.qmd
Submit Part 3

--- 2023-02-24 products/manuscript/Manuscript.qmd
@@
 ---
-title: "MADA Final Project"
-subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
-author: Nicole Luisi
+title: "Exploratory analysis script"
+author: "Nicole Luisi"
 date: "`r Sys.Date()`"
-format:
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../dataanalysis_template_references.bib
-csl: ../apa.csl
+output: html_document
 ---
 
-Note: There seem to be  lot of issues getting tables to render well in Word doc format, need to come back to this.
-I tried 5 or so different methods and they all look terrible when in a doc vs html.
-Also having issues with resolving crossrefs.
+# Setup
 
@@
 
-# Summary/Abstract
+## Load data
 
-The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years.
-Previous research by [Chapin-Bardales et al.
-(2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
-This analysis will focus on new HIV diagnosesbetween  2008-2022; subgroups will include region, age group, gender, and race/ethnicity.
-[@Chapin2017]
-
-# Introduction
-
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
-
-## Description of data and data source
-
-**Data Source**:
-
-*Note: New diagnoses data are currently available through 2020.
-If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
-
-+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]
-
-**Analysis**:
-
-+ Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
-+ Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
-+ Estimated annual percent change (EAPC) for incidence
-  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
-+ Other modeling TBD based on course modules
-
-## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
-
-{{< pagebreak >}}
-
-# Methods
-
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-
-## Data aquisition
-Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in the `data/raw_data/AIDSVu` subfolder.
-The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).
-
-## Data import and cleaning
-The `processingcode.R` file includes the following steps:
-+ Read in xlsx data files and combine all years for each level
-+ Subset variables of interest and rename variables so all levels are consistent 
-+ View/explore data
-+ Calculate new rate ratio variables for subgroups 
-+ Prepare combined national and regional file with all variables
-
-## Statistical analysis
-_Explain anything related to your statistical analyses._
-
-{{< pagebreak >}}
-
-# Results
+```{r}
+alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
+```
 
@@
 
-@tbl-natsummarytable shows a full summary of the National data.
+Full summary of the National data.
 
 ```{r}
-#| label: tbl-natsummarytable
-#| tbl-cap: "National data summary table."
-#| echo: FALSE
-alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
 national<-alldata %>% filter(Geo %in% c("United States"))
@@
 ```
-@fig-natplots shows exploratory plots for the national rates by subgroup.
+
+Exploratory plots for the national rates by subgroup.
 
 ```{r}
-#| label: fig-natplots
-#| fig-cap: "National rates by subgroup"
-#| echo: FALSE
 national<-alldata %>% filter(Geo %in% c("United States"))
@@
 
-@tbl-natratios shows a summary of selected rate ratios for race/ethnicity and sex.
+Summary of selected rate ratios for race/ethnicity and sex.
 
 ```{r}
-#| label: tbl-natratios
-#| tbl-cap: "National rate ratios by select subgroups"
-#| echo: FALSE
 national<-alldata %>% filter(Geo %in% c("United States"))
@@
 
-@tbl-regsummarytable shows a full summary of the Regional data.
+Full summary of the Regional data.
 
 ```{r}
-#| label: tbl-regsummarytable
-#| tbl-cap: "Regional data summary table."
-#| echo: FALSE
-alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
 regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
@@
 
-@fig-regplots shows an exploratory plot for regional overall new diagnosis rates with female-male rate ratios.
+Exploratory plot for regional overall new diagnosis rates with female-male rate ratios.
 
 ```{r}
-#| label: fig-regplots
-#| tbl-cap: "Regional plots"
-#| echo: FALSE
-alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
 regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
@@
 ```
-
-## Basic statistical analysis
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by sex."
-#| echo: FALSE
-#knitr::include_graphics(here("results","height_weight_stratified.png"))
-```
-
-## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-#resulttable2 = readRDS(here("results","resulttable2.rds"))
-#knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
-
-## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
-
-## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [] discusses types of analyses.
-
-These papers [] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
-{{< pagebreak >}}
-
-# References

### 2023-03-17 products/manuscript/Manuscript.qmd
Submit Part 3

--- 2023-02-03 products/manuscript/Manuscript.qmd
@@
 ---
-title: "MADA Final Project"
-subtitle: "Trends in PrEP Use in the US, 2012-2022"
+title: "MADA 8060E Project"
+subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
 author: Nicole Luisi
@@
 format:
-  docx:
+  html:
+    self-contained: true
     toc: false
@@
 
-# PART 1
-
-**Overview**:
-
-The purpose of this analysis is to measure the public health impact of pre-exposure prophylaxis (PrEP) in the US over the past 10 years.
-Previous research by [Sullivan et al.
-(2018)](https://doi.org/10.1016/j.annepidem.2018.06.009) evaluated trends in the prevalence of PrEP use from 2012-2017.
-Outcomes include period prevalence of PrEP use and PrEP-to-need ratio (PnR) by year.
-This analysis will extend the previous work to include data through the year 2022; subgroups will include region, state, age group, gender, and race/ethnicity *(race/ethnicity PrEP data were not available at the time of the previous analysis and were not included)*.
-[@SULLIVAN2018833]
-
-**Data Source**:
-
-*Note: New diagnoses data are currently available through 2020; PrEP data through 2021.
-If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
-
-+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020] 
-+ PrEP and PnR: IQVIA (Quintiles IMS Holdings)
-+ Annual Population Estimates: National Historical Geographic Information System (NHGIS), American Community Surveys (ACS), US Census Bureau
-
-**Analysis**:
-
-+ N/% of annual PrEP users by age, gender, race/ethnicity, and region
-+ Annual prevalence of PrEP use per 100,000 population by age, gender, race/ethnicity, region, and state
-+ Estimated annual percent change (EAPC) for prevalence
-  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
-+ Prep-to-need ratio by age, gender, race/ethnicity, and region
-+ Other modeling TBD based on course modules
-
-{{< pagebreak >}}
-
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-#library(here)
-#library(knitr)
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+# load R packages
+library(here) 
+library(ggplot2)
+library(skimr) 
+library(plotly)
+library(gt)
+library(knitr)
+library(readxl)
+library(dplyr)
+library(kableExtra)
+library(gridExtra)
+library(ggpubr)
 ```
@@
 # Summary/Abstract
-_Write a summary of your project._
 
-{{< pagebreak >}}
+The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years.
+Previous research by [Chapin-Bardales et al.
+(2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
+This analysis will focus on new HIV diagnoses between 2008-2022; subgroups will include region, age group, gender, and race/ethnicity.
+[@Chapin2017]
 
@@
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
+**Data Source**:
+
+*Note: New diagnoses data are currently only available through 2020 on AIDSVu and CDC Atlas.
+If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
+
+* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]
+
+**Analysis**:
+
+* Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
+* Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
+* Average annual percent change (AAPC) for incidence (calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation))
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in []
+This analysis will evaluate national and regional trends in HIV new diagnoses between 2008-2020 by sex, age, and race/ethnicity.
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in `data/raw_data/AIDSVu` sub-folders.
+The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+The `processingcode.R` file includes the following steps:
+
+* Read in xlsx data files and combine all years for each level
+* Subset variables of interest and rename variables so all levels are consistent 
+* View/explore data
+* Calculate new rate ratio variables for subgroups (White=ref)
+* Prepare combined national and regional file with all variables
+
+The `analysis.qmd` file includes the following steps:
+
+* Read in result data from external Joinpoint analyses
+* Clean and combine Joinpoint analysis results
+* View/explore data
+* Create tables and plots of results
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+Rate ratios for race/ethnicity were calculated by dividing the new diagnosis rate for each group by the new diagnosis rate for the reference group (non-Hispanic White).
+
+[Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation) was used to test for significant joinpoints (years) to identify  trends in new diagnoses rates, nationally and regionally, by sex, age, and race/ethnicity.
+Trends for rate ratios were also used to assess disparities by race/ethnicity over time.
+For Joinpoint analyses, the maximum number of joinpoints was set to 3, and standard errors were auto-calculated by the tool under the heteroscedasticity assumption.
+The average annual percent changes (AAPCs) from 2008-2020 were also calculated for all subsets.
 
@@
 
-## Exploratory/Descriptive analysis
+## Descriptive analysis
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+@fig-natplotsex shows line plot for national new diagnosis rates by sex.
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: fig-natplotsex
+#| fig-cap: "National new diagnosis rates by sex"
 #| echo: FALSE
-#resulttable=readRDS("../../results/summarytable.rds")
-#knitr::kable(resulttable)
+knitr::include_graphics(here("results","US_rates_bysex.png"))
 ```
 
-## Basic statistical analysis
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+@fig-natplotage shows line plot for national new diagnosis rates by age group.
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by sex."
+#| label: fig-natplotage
+#| fig-cap: "National new diagnosis rates by age group"
 #| echo: FALSE
-#knitr::include_graphics(here("results","height_weight_stratified.png"))
+knitr::include_graphics(here("results","US_rates_byage.png"))
 ```
 
-## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+@fig-natplotrace shows line plot for national new diagnosis rates by race/ethnicity.
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: fig-natplotrace
+#| fig-cap: "National new diagnosis rates by race/ethnicity"
 #| echo: FALSE
-#resulttable2 = readRDS(here("results","resulttable2.rds"))
-#knitr::kable(resulttable2)
+knitr::include_graphics(here("results","US_rates_byrace.png"))
+```
+
+Trends by sex and age group are as expected.
+For sex and age, all rates have declined between 2008-2020 and groups with the highest incidence had the steepest declines over time, particularly after 2016.
+The remainder of this analysis will focus on trends and disparities by race/ethnicity.
+
+@fig-regplotrace shows a bubble plot comparing regional new diagnosis (log) rates by race/ethnicity.
+The color represents race/ethnicity and the size of the point represents the log of the new diagnosis rate.
+
+```{r}
+#| label: fig-regplotrace
+#| fig-cap: "Regional new diagnosis rate (log) by race/ethnicity"
+#| echo: FALSE
+knitr::include_graphics(here("results","racerates_region.png"))
+```
+
+@fig-natplotraceratios shows national plot of log race ratios for new diagnoses (White=ref).
+
+```{r}
+#| label: fig-natplotraceratios
+#| fig-cap: "National race ratios"
+#| echo: FALSE
+knitr::include_graphics(here("results","racerates_allUS.png"))
+```
+
+@fig-natplotdiffbwh shows national plot of White-Black and White-Hispanic rate differences by year.
+
+```{r}
+#| label: fig-natplotdiffbwh
+#| fig-cap: "White-Black and White-Hispanic differences in national new diagnosis rate"
+#| echo: FALSE
+knitr::include_graphics(here("results","nat_diff_bwh.png"))
+```
+
+@tbl-allrates shows a full summary of new diagnosis rates for all US by group.
+
+```{r}
+#| label: tbl-allrates
+#| tbl-cap: "All US New diagnosis Rates"
+#| echo: FALSE
+resulttable=readRDS(here("results","all_us_rates.rds"))
+kable(resulttable,caption = "All US New Diagnosis Rates", col.names=c("Year", "Overall", "Male", "Female", "Black", "White", "Hispanic", "Asian", "American Indian/Alaska Native", "Multiracial", "Native Hawaiian/Pacific Islander", "Age 13-24", "Age 25-34", "Age 35-44", "Age 45-54", "Age 55+"), align="rc") %>%
+  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10)  
+```
+
+## Statistical analysis
+
+@tbl-allaapc shows a summary of average annual percent change (AAPC) for each subgroup and geographic region.
+
+```{r}
+#| label: tbl-allaapc
+#| tbl-cap: "All AAPCs"
+#| echo: FALSE
+resulttable=readRDS(here("results","aapc_all_results.rds"))
+kable(resulttable,caption = "All AAPC Results", 
+      col.names=c("Stratification", "Geographic Area", "AAPC", "AAPC_lCI", "AAPC_uCI", "p-value"),      align="rc") %>% 
+  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
+  pack_rows("Overall", 1, 5) %>%
+  pack_rows("Female", 6, 10) %>%
+  pack_rows("Male", 11, 15) %>%
+  pack_rows("Age 13-24", 16, 20) %>%
+  pack_rows("Age 25-34", 21, 25) %>%
+  pack_rows("Age 35-44", 26, 30)  %>%
+  pack_rows("Age 45-54", 31, 35)  %>%
+  pack_rows("Age 55+", 36, 40)  %>%
+  pack_rows("American Indian/Alaska Native", 41, 44) %>%
+  pack_rows("Asian", 45, 49) %>%
+  pack_rows("Black", 50, 54) %>%
+  pack_rows("Hispanic", 55, 59) %>%
+  pack_rows("Multiracial", 60, 64) %>%
+  pack_rows("Native Hawaiian/Pacific Islander", 65, 67) %>%
+  pack_rows("White", 68, 72) %>%
+  pack_rows("AIAN-White", 73, 76) %>%
+  pack_rows("Asian-White", 77, 81) %>%
+  pack_rows("Black-White", 82, 86) %>%
+  pack_rows("Hispanic-White", 87, 91) %>%
+  pack_rows("Multi-White", 92, 96) %>%
+  pack_rows("NHPI-White", 97, 99) 
+```
+
+@tbl-allapc shows a full summary of annual percent change (APC) for each segment of time within each subgroup and geographic region.
+
+```{r}
+#| label: tbl-allapc
+#| tbl-cap: "All APCs"
+#| echo: FALSE
+resulttable=readRDS(here("results","apc_all_results.rds"))
+kable(resulttable,caption = "All APC Results", align="rc") %>% 
+  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
+  pack_rows("Overall", 1, 15) %>%
+  pack_rows("Female", 16, 24) %>%
+  pack_rows("Male", 25, 32) %>%
+  pack_rows("Age 13-24", 33, 41) %>%
+  pack_rows("Age 25-34", 42, 54) %>%
+  pack_rows("Age 35-44", 55, 63)  %>%
+  pack_rows("Age 45-54", 64, 70)  %>%
+  pack_rows("Age 55+", 71, 76)  %>%
+  pack_rows("American Indian/Alaska Native", 77, 80) %>%
+  pack_rows("Asian", 81, 90) %>%
+  pack_rows("Black", 91, 102) %>%
+  pack_rows("Hispanic", 103, 113) %>%
+  pack_rows("Multiracial", 114, 120) %>%
+  pack_rows("Native Hawaiian/Pacific Islander", 121, 123) %>%
+  pack_rows("White", 124, 130) %>%
+  pack_rows("AIAN-White", 131, 134) %>%
+  pack_rows("Asian-White", 135, 142) %>%
+  pack_rows("Black-White", 143, 147) %>%
+  pack_rows("Hispanic-White", 148, 156) %>%
+  pack_rows("Multi-White", 157, 165) %>%
+  pack_rows("NHPI-White", 166, 168) 
+```
+
+@fig-natplotbwhjpoint shows national plot of new diagnosis Joinpoints for race/ethnicity (Black, White, Hispanic).
+
+```{r}
+#| label: fig-natplotbwhjpoint
+#| fig-cap: "National Joinpoint results comparing Black, White, Hispanic"
+#| echo: FALSE
+knitr::include_graphics(here("results","raceBWH_joinpoints_allUS.png"))
+```
+
+@fig-natplotraceratioaapc shows regional plot of new diagnosis AAPCs for race/ethnicity.
+
+```{r}
+#| label: fig-natplotraceratioaapc
+#| fig-cap: "Regional Joinpoint AAPC results by race/ethnicity"
+#| echo: FALSE
+knitr::include_graphics(here("results","raceratio_byregion.png"))
 ```
@@
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [] discusses types of analyses.
-
-These papers [] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
-
 {{< pagebreak >}}
 
-# References
+## References

### 2023-03-17 code/analysis_code/joinpoint_analysis.R

@@ -1,0 +1,194 @@
+###############################
+# Joinpoint analysis script
+#
+# This script loads the processed, cleaned data, does some analysis and saves the results 
+
+######################################
+######################################
+
+# NOTE: 
+
+# This analysis uses a proprietary software package available from the NIH (https://surveillance.cancer.gov/joinpoint/)
+# The R callable software must be requested from NIH and installed on the users local machine
+# The nih.joinpoint package referenced here will NOT work unless the R callable version of the software is installed 
+
+# Because of this and some issues with specific R callable options, the desktop version of the software was used for some analyses
+# All result tables from external joinpoint analyses are available in the data/processed_data folder so tables and plots can be reproduced 
+
+######################################
+######################################
+
+# Load packages
+library(ggplot2) 
+library(kableExtra)
+library(here)
+library(tidyverse)
+#devtools::install_github("DanChaltiel/nih.joinpoint")
+library(nih.joinpoint)
+
+# Path to data using HERE package
+data_location <- here::here("data","processed_data","processeddata.rds")
+
+######################################
+# Data fitting/statistical analysis
+######################################
+
+# Joinpoint program settings/options
+options(joinpoint_path="C:/Program Files (x86)/Joinpoint Command/jpCommand.exe")
+run_opt = run_options(model="ln", max_joinpoints=3, n_cores=3)
+export_opt = export_options()
+
+
+# Overall APC
+jp_tot_run = joinpoint(jp_tot, x=Year, y=Rate, by=Geo,
+                       run_opts=run_opt, export_opts=export_opt)
+jp_total_results<-jp_tot_run[["apc"]]
+
+jp_total_results<-jp_total_results %>% dplyr::select(-c("model", "segment", "apc_significant"))
+jp_total_results<-jp_total_results %>% arrange(match(geo, c("United States", "Northeast", "Midwest", "South", "West")), (segment_start))
+kable(jp_total_results,caption = "Joinpoint Results, Overall by Geographic Region", 
+      col.names=c("Region", "Start Year", "End Year", "Annual % Change (APC)", "APC 95LCL", "APC 95UCL", 
+                  "Test Statistic",  "p-value"), align="rc") %>% 
[…]

### 2023-03-17 code/processing_code/processingcode.R

@@ -3,11 +3,6 @@
 #
 #this script loads the raw data, processes and cleans it 
 #and saves it as Rds file in the processed_data folder
-#
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
-
 
 ## ---- packages --------
 #load needed packages. make sure they are installed.
@@ -18,6 +13,7 @@ library(skimr) #for nice visualization of data
 library(here) #to set paths
 library(plotly)
 library(gt)
+library(reshape2)
 
 ## ---- loaddata --------
 #path to data
@@ -89,7 +85,7 @@ national_sub_newdx$nhpi_white_rateratio<-round(national_sub_newdx$NHPI_Rate/nati
 national_sub_newdx$age13t24_age35t44_rateratio<-round(national_sub_newdx$Age13t24_Rate/national_sub_newdx$Age35t44_Rate, digits=1)
 national_sub_newdx$age25t34_age35t44_rateratio<-round(national_sub_newdx$Age25t34_Rate/national_sub_newdx$Age35t44_Rate, digits=1)
 national_sub_newdx$age45t54_age35t44_rateratio<-round(national_sub_newdx$Age45t54_Rate/national_sub_newdx$Age35t44_Rate, digits=1)
-national_sub_newdx$age45t54_age55p_rateratio<-round(national_sub_newdx$Age55p_Rate/national_sub_newdx$Age35t44_Rate, digits=1)
+national_sub_newdx$age55p_age35t44_rateratio<-round(national_sub_newdx$Age55p_Rate/national_sub_newdx$Age35t44_Rate, digits=1)
 
 # regional
 # create rate ratios and round 
@@ -119,7 +115,76 @@ allnewdx<-rbind(national_sub_newdx,regional_sub_newdx)
 save_data_location <- here::here("data","processed_data","allnewdx.rds")
 saveRDS(allnewdx, file = save_data_location)
 
+## ---- prepare file formatted for joinpoint program --------
+alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
+names(alldata)
 
+# Race
+jp_race<-melt(alldata, id.vars = c("Year", "Geo"),
[…]

### 2023-04-07 code/analysis_code/analysis.qmd
Submit Part 4

@@
 library(janitor)
+library(tidymodels)
+library(dotwhisker)
+library(tidyverse)
+library(caret)
 ```
@@
 ```
+
+# Analyze State Race Data with Google Trend HIV Search
+
+```{r}
+state_race_gtrend<-readRDS(here("data", "processed_data", "state_race_gtrend.rds"))
+gtrends_2021_2022 <- readxl::read_excel(here("data","raw_data", "Misc", "Gtrends_2021_2022.xlsx")) 
+p_2022 <- gtrends_2021_2022 %>% filter(Year %in% c(2022))
+
+linear_reg() %>% 
+  set_engine("keras")
+lm_mod <- linear_reg()
+
+# predict 2022 black-white
+lm_fit <- 
+  lm_mod %>% 
+  fit(black_white_rateratio ~ Gtrend + Year + State , data = state_race_gtrend)
+lm_fit
+tidy(lm_fit)
+tidy(lm_fit) %>% 
+  dwplot(dot_args = list(size = 2, color = "black"),
+         whisker_args = list(color = "black"),
+         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
+
+mean_pred <- predict(lm_fit, new_data = p_2022)
+conf_int_pred <- predict(lm_fit, new_data = p_2022, type = "conf_int")
+glance(lm_fit) %>%
+  dplyr::select(adj.r.squared, sigma, AIC, BIC, p.value)
+
+plot_data <- 
+  p_2022 %>% 
+  bind_cols(mean_pred) %>% 
+  bind_cols(conf_int_pred)
+
+compare <- state_race_gtrend  %>% filter(Year %in% c(2020)) %>% select(State, black_white_rateratio, hispanic_white_rateratio) 
+plot_data_c <- merge(plot_data,compare,by=c("State"),all=T)
+
+south<-plot_data_c %>% filter(State %in% c('Alabama','Arkansas','Delaware','District of Columbia','Florida','Georgia',
+                                         'Kentucky','Louisiana','Maryland','Mississippi','North Carolina','Oklahoma',
+                                         'South Carolina','Tennessee','Texas','Virginia','West Virginia'))
+northeast<-plot_data_c %>% filter(State %in% c('Connecticut', 'Maine', 'Massachusetts', 'New Hampshire', 'New Jersey', 'New York', 
+                                             'Pennsylvania', 'Rhode Island', 'Vermont'))
+west<-plot_data_c %>% filter(State %in% c('Alaska', 'Arizona', 'California', 'Colorado', 'Hawaii', 'Idaho', 'Montana', 'Nevada', 'New Mexico',
+                                        'Oregon', 'Utah', 'Washington', 'Wyoming'))
+midwest<-plot_data_c %>% filter(State %in% c('Illinois', 'Indiana', 'Iowa', 'Kansas', 'Michigan', 'Minnesota', 'Missouri', 'Nebraska', 'North Dakota',
+                                           'Ohio','South Dakota', 'Wisconsin'))
+
+south22<-ggplot(south, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Black-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('AL','AR','DE','DC','FL','GA', 'KY','LA','MD','MS','NC','OK',
+                                         'SC','TN','TX','VA','WV')) +
+  ggtitle(label="SOUTH: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = black_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))
+
+northeast22<-ggplot(northeast, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Black-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('CT', 'ME', 'MA', 'NH', 'NJ', 'NY', 'PA', 'RI', 'VT')) +
+  ggtitle(label="NORTHEAST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = black_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))
+
+west22<-ggplot(west, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Black-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('AK', 'AZ', 'CA', 'CO', 'HI', 'ID', 'MT', 'NV', 'NM', 
+                            'OR', 'UT', 'WA', 'WY')) +
+  ggtitle(label="WEST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = black_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))
+
+midwest22<-ggplot(midwest, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Black-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('IL', 'IN', 'IA', 'KS', 'MI', 'MN', 'MO', 'NE', 'ND',
+                            'OH','SD', 'WI')) +
+  ggtitle(label="MIDWEST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = black_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))
+
+predict_2022_blackwhite<-ggarrange(northeast22, south22, west22, midwest22, ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")
+predict_2022_blackwhite
+figure_file = here("results","Predict2022_BWratio.png")
+ggsave(filename = figure_file, plot=predict_2022_blackwhite)
+
+# predict 2022 hispanic-white
+lm_fit <- 
+  lm_mod %>% 
+  fit(hispanic_white_rateratio ~ Gtrend + Year + State , data = state_race_gtrend)
+lm_fit
+tidy(lm_fit)
+tidy(lm_fit) %>% 
+  dwplot(dot_args = list(size = 2, color = "black"),
+         whisker_args = list(color = "black"),
+         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2)) 
+mean_pred <- predict(lm_fit, new_data = p_2022)
+conf_int_pred <- predict(lm_fit, new_data = p_2022, type = "conf_int")
+glance(lm_fit) %>%
+  dplyr::select(adj.r.squared, sigma, AIC, BIC, p.value)
+
+plot_data <- 
+  p_2022 %>% 
+  bind_cols(mean_pred) %>% 
+  bind_cols(conf_int_pred)
+
+compare <- state_race_gtrend  %>% filter(Year %in% c(2020)) %>% select(State, black_white_rateratio, hispanic_white_rateratio) 
+plot_data_c <- merge(plot_data,compare,by=c("State"),all=T)
+
+south<-plot_data_c %>% filter(State %in% c('Alabama','Arkansas','Delaware','District of Columbia','Florida','Georgia',
+                                         'Kentucky','Louisiana','Maryland','Mississippi','North Carolina','Oklahoma',
+                                         'South Carolina','Tennessee','Texas','Virginia','West Virginia'))
+northeast<-plot_data_c %>% filter(State %in% c('Connecticut', 'Maine', 'Massachusetts', 'New Hampshire', 'New Jersey', 'New York', 
+                                             'Pennsylvania', 'Rhode Island', 'Vermont'))
+west<-plot_data_c %>% filter(State %in% c('Alaska', 'Arizona', 'California', 'Colorado', 'Hawaii', 'Idaho', 'Montana', 'Nevada', 'New Mexico',
+                                        'Oregon', 'Utah', 'Washington', 'Wyoming'))
+midwest<-plot_data_c %>% filter(State %in% c('Illinois', 'Indiana', 'Iowa', 'Kansas', 'Michigan', 'Minnesota', 'Missouri', 'Nebraska', 'North Dakota',
+                                           'Ohio','South Dakota', 'Wisconsin'))
+
+south22<-ggplot(south, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Hispanic-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('AL','AR','DE','DC','FL','GA', 'KY','LA','MD','MS','NC','OK',
+                                         'SC','TN','TX','VA','WV')) +
+  ggtitle(label="SOUTH: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))
+
+northeast22<-ggplot(northeast, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Hispanic-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('CT', 'ME', 'MA', 'NH', 'NJ', 'NY', 'PA', 'RI', 'VT')) +
+  ggtitle(label="NORTHEAST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))
+
+west22<-ggplot(west, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Hispanic-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('AK', 'AZ', 'CA', 'CO', 'HI', 'ID', 'MT', 'NV', 'NM', 
+                            'OR', 'UT', 'WA', 'WY')) +
+  ggtitle(label="WEST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))
+
+midwest22<-ggplot(midwest, aes(x = State)) + 
+  geom_point(aes(y = .pred)) + 
+  geom_errorbar(aes(ymin = .pred_lower, 
+                    ymax = .pred_upper),
+                width = .2) + 
+  labs(y = "Hispanic-White Rate-Ratio") + 
+  scale_x_discrete(labels=c('IL', 'IN', 'IA', 'KS', 'MI', 'MN', 'MO', 'NE', 'ND',
+                            'OH','SD', 'WI')) +
+  ggtitle(label="MIDWEST: Predict 2022") + 
+  theme_light() + 
+  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
+  theme(legend.title = element_blank()) + 
+  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))
+
+predict_2022_hispanicwhite<-ggarrange(northeast22, south22, west22, midwest22, ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")
+predict_2022_hispanicwhite
+figure_file = here("results","Predict2022_HWratio.png")
+ggsave(filename = figure_file, plot=predict_2022_hispanicwhite) 
+```

### 2023-04-07 products/manuscript/Manuscript.qmd
Submit Part 4

@@
 title: "MADA 8060E Project"
-subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
+subtitle: "Trends in New HIV Diagnoses in the US, 2008-2020"
 author: Nicole Luisi
@@
 
-The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years.
-Previous research by [Chapin-Bardales et al.
-(2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
-This analysis will focus on new HIV diagnoses between 2008-2022; subgroups will include region, age group, gender, and race/ethnicity.
-[@Chapin2017]
+The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years (2008-2020) by region, age group, gender, and race/ethnicity.
+Joinpoint regression was used to identify significant trends and annual percent changes in new diagnoses rates, and trends for rate ratios were also used to assess disparities by race/ethnicity over time.
+Google Trend data from 2016-2020 for search term "HIV" was used to predict not yet released Black-White and Hispanic-White rate ratios for 2022.
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Previous research by [Chapin-Bardales et al. (2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
+
+[@Chapin2017] [@Sullivan2020] [@SULLIVAN2018833] [@FIELDS2021] [@BOSH2021] [@MCCREE2019] [@YOUNG2018]
 
@@
 
-* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]
+* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) 
+* Google Trend data for search term "HIV": [Google Trends](https://trends.google.com/)
+
+[@Sullivan2020]
 
@@
 * Average annual percent change (AAPC) for incidence (calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation))
+* Predictions for Black-White and Hispanic-White rate ratios by state, using Google Trend search data for the term "HIV"
 
@@
 This analysis will evaluate national and regional trends in HIV new diagnoses between 2008-2020 by sex, age, and race/ethnicity.
+Black-White and Hispanic-White rate ratios will also be predicted for 2022 based on Google Trend results for search term "HIV".
 
@@
 Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in `data/raw_data/AIDSVu` sub-folders.
-The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).
+The new diagnosis files consist of one xlsx file per year for each level (national, regional, state, etc.).
+For [Google Trends](https://trends.google.com/), annual estimates for search term "HIV" were downloaded for each state.
 
@@
 The average annual percent changes (AAPCs) from 2008-2020 were also calculated for all subsets.
+
+Google Trend data for search term "HIV" was downloaded for each state and year between 2016-2020.
+These data were used in linear models to predict not yet released 2022 Black-White and Hispanic-White rate ratios based on Google Trend data for 2022.
 
@@
 
+@fig-bwrrpredict22 shows state-level predictions for Black-White Rate Ratios in 2022 based on Google Trend data for Search Term "HIV".
+
+```{r}
+#| label: fig-bwrrpredict22
+#| fig-cap: "Black-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
+#| echo: FALSE
+knitr::include_graphics(here("results","Predict2022_BWratio.png"))
+```
+
+@fig-hwrrpredict22 shows state-level predictions for Hispanic-White Rate Ratios in 2022 based on Google Trend data for Search Term "HIV".
+
+```{r}
+#| label: fig-hwrrpredict22
+#| fig-cap: "Hispanic-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
+#| echo: FALSE
+knitr::include_graphics(here("results","Predict2022_HWratio.png"))
+```
+
 {{< pagebreak >}}

### 2023-04-07 code/processing_code/processingcode.R

@@ -20,22 +20,6 @@ library(reshape2)
 #note the use of the here() package and not absolute paths
 nat_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows  
 reg_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "Region", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#st_newdx_files <- list.files(path = (here("data","raw_data", "AIDSVu", "State", "NewDx")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-
-#nat_pnr_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "PNR")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#nat_prep_files <- list.files(path = (here("data","raw_data", "AIDSVu", "National", "PrEP")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#reg_pnr_files <- list.files(path = (here("data","raw_data", "AIDSVu", "Region", "PNR")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#reg_prep_files <- list.files(path = (here("data","raw_data", "AIDSVu", "Region", "PrEP")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#st_pnr_files <- list.files(path = (here("data","raw_data", "AIDSVu", "State", "PNR")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-#st_prep_files <- list.files(path = (here("data","raw_data", "AIDSVu", "State", "PrEP")), pattern = "*.xlsx", full.names = T) %>% lapply(readxl::read_excel, skip = 2, col_names = T) %>% bind_rows 
-
[…]

### 2023-04-21 code/analysis_code/analysis.qmd
Submit Part 5

@@
 natfig_a <- natfig_a %>% add_trace(y = ~Age25t34_Rate, name = "Age 25-34", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
-natfig_a <- natfig_a %>% add_trace(y = ~Age35t44_Rate, name = "Age 35-44e", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
+natfig_a <- natfig_a %>% add_trace(y = ~Age35t44_Rate, name = "Age 35-44", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
 natfig_a <- natfig_a %>% add_trace(y = ~Age45t54_Rate, name = "Age 45-54", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))

### 2023-04-21 code/analysis_code/exploratory_analysis.qmd
Submit Part 5

@@
 library(knitr)
+library(dplyr)
 ```

### 2023-04-21 products/manuscript/Manuscript.qmd
Submit Part 5

--- 2023-03-17 products/manuscript/Manuscript.qmd
@@
 format:
-  html:
-    self-contained: true
+  docx:
     toc: false
@@
     highlight-style: github
+    prefer-html: true
 bibliography: ../dataanalysis_template_references.bib
@@
 
-The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years.
-Previous research by [Chapin-Bardales et al.
-(2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.
-This analysis will focus on new HIV diagnoses between 2008-2022; subgroups will include region, age group, gender, and race/ethnicity.
-[@Chapin2017]
+This report evaluates trends in new HIV diagnoses over the past 10 years (2008-2020) by region, age group, gender, and race/ethnicity.
+Joinpoint regression was used to identify significant trends and annual percent changes in new diagnoses rates, and trends for rate ratios were also used to assess disparities by race/ethnicity over time.
+Google Trend data from 2016-2020 for search term "HIV" was used to predict unreleased Black-White and Hispanic-White rate ratios for 2022.
+Overall, new HIV diagnoses declined for all groups over the last several years.
+Race ratios are predicted to continue to decline within most US states.
+More data is needed to determine whether the steepest declines observed between 2019-2020 are due to programmatic changes or other issues related to the COVID-19 pandemic.
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Over the last 15 years, the United States (US) has made substantial progress in reducing the number of new HIV diagnoses each year.
+[@MILLETT2010, @BOSH2021] A number of effective programmatic and clinical-based strategies to prevent HIV transmission have led to a steady decline in new HIV diagnoses rates for several years, but for many groups this trend began to wane around 2013, prompting new initiatives such as ['Ending the HIV Epidemic'](https://ahead.hiv.gov/), a national plan to increase the use of a number of key prevention strategies and to reach the goal of a 90% reduction in new diagnoses by 2023.
+[@CDCStateEpid22, @HIVGOV19] In recent years, new HIV diagnoses rates have continued to decline, but a number of socio-demographic and geographic disparities still remain [@CDCStateEpid22], as does some uncertainty about the impact of current prevention strategies [@BRADLEY2019] and the COVID-19 epidemic, particularly among vulnerable or high risk groups [@CDCStateEpid22, @FIELDS2021, @MCCREE2019].
+Previous research has evaluated trends in new AIDS diagnosis rates by race/ethnicity between 1984-2013 [@Chapin2017], and trends in the use of pre-exposure prophylaxis (PrEP) between 2007-2016 [@SULLIVAN2018833].
+Using similar methods, this project evaluates trends for new HIV diagnoses by race/ethnicity and other subgroups between 2008-2020, and a sub-analysis inspired by Young et al.
+(2018) [@YOUNG2018] uses publicly available [Google Trends](https://trends.google.com/) data to predict state-level new HIV diagnoses rates for 2022, which have not yet been released.
 
@@
 
-*Note: New diagnoses data are currently only available through 2020 on AIDSVu and CDC Atlas.
-If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*
+New HIV diagnoses data at the national, regional, and state levels were obtained from [AIDSVu](https://aidsvu.org/), and interactive online tool developed by researchers at Emory University that allows users to explore HIV-related data compiled from the Centers for Disease Control and Prevention (CDC) and other sources.
+[@Sullivan2020] *New diagnoses data were only available through 2020 at the time of this report.*
 
-* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]
+The [Google Trends](https://trends.google.com/) website was used to extract state-level data for the Google search term "HIV" in 2022.
+This site allows users to explore and download data on the use Google search terms, stratified by a number of different factors.
 
@@
 
-* Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
-* Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
-* Average annual percent change (AAPC) for incidence (calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation))
+Data from [AIDSVu](https://aidsvu.org/) were compiled for the variables of interest (new diagnoses rates overall and by sex, age, and race/ethnicity), at the national, regional, and state level.
+Detailed data methods are available on the [AIDSVu website](https://aidsvu.org/data-methods/), but briefly, new diagnoses include anyone with an HIV diagnosis in each 1-year period; all rates were calculated per 100,000.
+For race/ethnicity comparisons, rate ratios were calculated using White non-Hispanic as the reference group.
 
-## Questions/Hypotheses to be addressed
-This analysis will evaluate national and regional trends in HIV new diagnoses between 2008-2020 by sex, age, and race/ethnicity.
+Annual percent change (APC) and average APC (AAPC) for incidence was calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)).
+These methods allow for a summary and trend analysis of rates within fixed periods of time.
+
+New diagnoses rate trends were explored for all geographic levels and stratification groups, and select results were summarized in this report.
+Based on previous research and results of the exploratory analysis, an emphasis was placed on evaluating trends by race/ethnicity.
+
+To predict 2022 Black-White and Hispanic-White new diagnoses rate ratios not yet released at the time of this report, [Google Trends](https://trends.google.com/) data for the search term "HIV" was included in a linear prediction model.
 
@@
 Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in `data/raw_data/AIDSVu` sub-folders.
-The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).
+The new diagnoses files consist of one xlsx file per year for each level (national, regional, state, etc.).
+For [Google Trends](https://trends.google.com/), annual estimates for search term "HIV" were downloaded for each state.
 
@@
 * View/explore data
+* Generate race ratio predictions for 2022
 * Create tables and plots of results
@@
 ## Statistical analysis
-Rate ratios for race/ethnicity were calculated by dividing the new diagnosis rate for each group by the new diagnosis rate for the reference group (non-Hispanic White).
+Rate ratios for race/ethnicity were calculated by dividing the new diagnosis rate for each sub-group by the new diagnosis rate for the reference group (non-Hispanic White).
 
@@
 The average annual percent changes (AAPCs) from 2008-2020 were also calculated for all subsets.
+
+Google Trend data for the search term "HIV" were downloaded for each state and year between 2016-2020.
+These data were used in linear models to predict not yet released 2022 Black-White and Hispanic-White rate ratios.
+`Tidymodels` was used to create a linear prediction model and metrics to evaluate model fit.
 
@@
 
-@fig-natplotsex shows line plot for national new diagnosis rates by sex.
-
-```{r}
-#| label: fig-natplotsex
-#| fig-cap: "National new diagnosis rates by sex"
-#| echo: FALSE
-knitr::include_graphics(here("results","US_rates_bysex.png"))
-```
-
-@fig-natplotage shows line plot for national new diagnosis rates by age group.
-
-```{r}
-#| label: fig-natplotage
-#| fig-cap: "National new diagnosis rates by age group"
-#| echo: FALSE
-knitr::include_graphics(here("results","US_rates_byage.png"))
-```
-
-@fig-natplotrace shows line plot for national new diagnosis rates by race/ethnicity.
-
-```{r}
-#| label: fig-natplotrace
-#| fig-cap: "National new diagnosis rates by race/ethnicity"
-#| echo: FALSE
-knitr::include_graphics(here("results","US_rates_byrace.png"))
-```
-
-Trends by sex and age group are as expected.
-For sex and age, all rates have declined between 2008-2020 and groups with the highest incidence had the steepest declines over time, particularly after 2016.
-The remainder of this analysis will focus on trends and disparities by race/ethnicity.
-
-@fig-regplotrace shows a bubble plot comparing regional new diagnosis (log) rates by race/ethnicity.
-The color represents race/ethnicity and the size of the point represents the log of the new diagnosis rate.
-
-```{r}
-#| label: fig-regplotrace
-#| fig-cap: "Regional new diagnosis rate (log) by race/ethnicity"
-#| echo: FALSE
-knitr::include_graphics(here("results","racerates_region.png"))
-```
-
-@fig-natplotraceratios shows national plot of log race ratios for new diagnoses (White=ref).
-
-```{r}
-#| label: fig-natplotraceratios
-#| fig-cap: "National race ratios"
-#| echo: FALSE
-knitr::include_graphics(here("results","racerates_allUS.png"))
-```
-
-@fig-natplotdiffbwh shows national plot of White-Black and White-Hispanic rate differences by year.
-
-```{r}
-#| label: fig-natplotdiffbwh
-#| fig-cap: "White-Black and White-Hispanic differences in national new diagnosis rate"
-#| echo: FALSE
-knitr::include_graphics(here("results","nat_diff_bwh.png"))
-```
-
-@tbl-allrates shows a full summary of new diagnosis rates for all US by group.
+The rate of new HIV diagnoses in the US has declined each year between 2008-2020, and in 2020 the US observed its lowest rate of 10.9 per 100,000.
+@tbl-allrates displays national new diagnosis rates overall, and by sex, age, and race/ethnicity.
+Similar to the overall trend observed, new diagnoses rates declined for both males and females during this period.
+@fig-natplotsex uses a line plot to illustrate national new diagnoses rates by sex.
+Rates were consistently higher for males versus females each year, and the decrease between 2008-2020 was twice that for males than females.
+@fig-natplotage uses a line plot to illustrate national new diagnoses rates by age group.
+Rates for all age groups also decreased during this time, with the sharpest decline among those age 25-34 between 2016-2019.
 
@@
 
+```{r}
+#| label: fig-natplotsex
+#| fig-cap: "National new diagnosis rates by sex"
+#| echo: FALSE
+knitr::include_graphics(here("results","US_rates_bysex.png"))
+```
+
+```{r}
+#| label: fig-natplotage
+#| fig-cap: "National new diagnosis rates by age group"
+#| echo: FALSE
+knitr::include_graphics(here("results","US_rates_byage.png"))
+```
+
+@fig-natplotrace uses a line plot to illustrate national new diagnoses rates by race/ethnicity.
+Rates were stable for most groups over this time period, with the exception of Multiracial and Black non-Hispanic persons; both of these groups showed a steep decline, particularly between 2019-2020.
+@fig-natplotracerates and @fig-regplotrace use bubble plots to compare new diagnoses (log) rates by race/ethnicity, nationally and regionally.
+In these plots, the color represents race/ethnicity and the size of the point represents the log of the new diagnoses rates.
+At the national level, consistent with previous research, rates for White non-Hispanic and Asian persons have remained relatively stable, while Black non-Hispanic, Multiracial, and Hispanic persons experience higher rates overall and therefore have some of the steepest declines in recent years.
+At the regional level, while overall trends are similar, larger disparities between groups were observed in the South.
+
+```{r}
+#| label: fig-natplotrace
+#| fig-cap: "National new diagnosis rates by race/ethnicity"
+#| echo: FALSE
+knitr::include_graphics(here("results","US_rates_byrace.png"))
+```
+
+```{r}
+#| label: fig-natplotracerates
+#| fig-cap: "National race ratios"
+#| echo: FALSE
+knitr::include_graphics(here("results","racerates_allUS.png"))
+```
+
+```{r}
+#| label: fig-regplotrace
+#| fig-cap: "Regional new diagnosis rate (log) by race/ethnicity"
+#| echo: FALSE
+knitr::include_graphics(here("results","racerates_region.png"))
+```
+
+Rates for Black non-Hispanic and Hispanic persons were evaluated using comparisons with a reference group (White non-Hispanic).
+Black-White and Hispanic-White rate ratios were 8.1 and 3.8 in 2020 (respectively), down from 9.7 and 4.3 in 2008.
+@fig-natplotdiffbwh displays national White-Black and White-Hispanic rate differences by year.
+
+```{r}
+#| label: fig-natplotdiffbwh
+#| fig-cap: "White-Black and White-Hispanic differences in national new diagnosis rate"
+#| echo: FALSE
+knitr::include_graphics(here("results","nat_diff_bwh.png"))
+```
+
 ## Statistical analysis
 
-@tbl-allaapc shows a summary of average annual percent change (AAPC) for each subgroup and geographic region.
+Tables @tbl-allapc and @tbl-allaapc summarize annual percent change (APC) and average annual percent change (AAPC) for each subgroup and geographic region.
+Race/ethnicity is of particular interest, so the remainder of the Joinpoint results will be within this subgroup.
+
+```{r}
+#| label: tbl-allapc
+#| tbl-cap: "All APCs"
+#| echo: FALSE
+resulttable=readRDS(here("results","apc_all_results.rds"))
+kable(resulttable,caption = "All APC Results", align="rc") %>% 
+  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
+  pack_rows("Overall", 1, 15) %>%
+  pack_rows("Female", 16, 24) %>%
+  pack_rows("Male", 25, 32) %>%
+  pack_rows("Age 13-24", 33, 41) %>%
+  pack_rows("Age 25-34", 42, 54) %>%
+  pack_rows("Age 35-44", 55, 63)  %>%
+  pack_rows("Age 45-54", 64, 70)  %>%
+  pack_rows("Age 55+", 71, 76)  %>%
+  pack_rows("American Indian/Alaska Native", 77, 80) %>%
+  pack_rows("Asian", 81, 90) %>%
+  pack_rows("Black", 91, 102) %>%
+  pack_rows("Hispanic", 103, 113) %>%
+  pack_rows("Multiracial", 114, 120) %>%
+  pack_rows("Native Hawaiian/Pacific Islander", 121, 123) %>%
+  pack_rows("White", 124, 130) %>%
+  pack_rows("AIAN-White", 131, 134) %>%
+  pack_rows("Asian-White", 135, 142) %>%
+  pack_rows("Black-White", 143, 147) %>%
+  pack_rows("Hispanic-White", 148, 156) %>%
+  pack_rows("Multi-White", 157, 165) %>%
+  pack_rows("NHPI-White", 166, 168) 
+```
 
@@
 
-@tbl-allapc shows a full summary of annual percent change (APC) for each segment of time within each subgroup and geographic region.
-
-```{r}
-#| label: tbl-allapc
-#| tbl-cap: "All APCs"
-#| echo: FALSE
-resulttable=readRDS(here("results","apc_all_results.rds"))
-kable(resulttable,caption = "All APC Results", align="rc") %>% 
-  row_spec(row = 0, color = "white", background = "#004987") %>% kable_styling(full_width=T, bootstrap_options = "striped", font_size=10) %>% 
-  pack_rows("Overall", 1, 15) %>%
-  pack_rows("Female", 16, 24) %>%
-  pack_rows("Male", 25, 32) %>%
-  pack_rows("Age 13-24", 33, 41) %>%
-  pack_rows("Age 25-34", 42, 54) %>%
-  pack_rows("Age 35-44", 55, 63)  %>%
-  pack_rows("Age 45-54", 64, 70)  %>%
-  pack_rows("Age 55+", 71, 76)  %>%
-  pack_rows("American Indian/Alaska Native", 77, 80) %>%
-  pack_rows("Asian", 81, 90) %>%
-  pack_rows("Black", 91, 102) %>%
-  pack_rows("Hispanic", 103, 113) %>%
-  pack_rows("Multiracial", 114, 120) %>%
-  pack_rows("Native Hawaiian/Pacific Islander", 121, 123) %>%
-  pack_rows("White", 124, 130) %>%
-  pack_rows("AIAN-White", 131, 134) %>%
-  pack_rows("Asian-White", 135, 142) %>%
-  pack_rows("Black-White", 143, 147) %>%
-  pack_rows("Hispanic-White", 148, 156) %>%
-  pack_rows("Multi-White", 157, 165) %>%
-  pack_rows("NHPI-White", 166, 168) 
-```
-
-@fig-natplotbwhjpoint shows national plot of new diagnosis Joinpoints for race/ethnicity (Black, White, Hispanic).
+National Joinpoints for race/ethnicity (Black non-Hispanic, White non-Hispanic, Hispanic) are summarized in @fig-natplotbwhjpoint.
+Of note, White non-Hispanic remain fairy stable, while disparities are evident for Black non-Hispanic and Hispanic persons, although rates show a steady decline, particularly after 2016.
+@fig-regplotraceratioaapc displays AAPC plots by region and race/ethnicity with 95% confidence intervals (CIs).
+AAPC was similar for Black non-Hispanic and Hispanic persons by region, and was most notable for Multiracial persons in all regions consistent with the steep decline observed in previous output.
 
@@
 
-@fig-natplotraceratioaapc shows regional plot of new diagnosis AAPCs for race/ethnicity.
-
 ```{r}
-#| label: fig-natplotraceratioaapc
+#| label: fig-regplotraceratioaapc
 #| fig-cap: "Regional Joinpoint AAPC results by race/ethnicity"
@@
 knitr::include_graphics(here("results","raceratio_byregion.png"))
+```
+
+For 2022 predictions, state-level data has been grouped and presented by US region.
+@fig-bwrrpredict22 and @fig-hwrrpredict22 display state-level predictions for Black-White and Hispanic-White rate ratios in 2022, based on the popularity of Google Trends search data for the search term "HIV".
+In these plots, the actual rate ratio observed for each state in 2020 is displayed in red, and the predicted rates and 95% CIs are in black.
+In most states, the observed 2020 rate ratio falls within the predicted range, with some exceptions.
+In Wyoming, the search term data predicted a much different Black-White rate ratio than previously observed, and a similar issue was noted in Vermont for the Hispanic-White ratio.
+Performance in this limited model was poor, with an adjusted R-squared of 0.55.
+
+```{r}
+#| label: fig-bwrrpredict22
+#| fig-cap: "Black-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
+#| echo: FALSE
+knitr::include_graphics(here("results","Predict2022_BWratio.png"))
+```
+
+```{r}
+#| label: fig-hwrrpredict22
+#| fig-cap: "Hispanic-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
+#| echo: FALSE
+knitr::include_graphics(here("results","Predict2022_HWratio.png"))
 ```
@@
 
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+This report summarizes general trends in new HIV diagnoses between 2008-2020.
+Overall, these trends were as expected, with a general overall decline for all groups over the last several years.
+The steepest decline in rates was observed between 2019-2020, among Black non-Hispanic and Multiracial persons.
+This decline may be due to programmatic changes or reporting issues during the COVID-19 pandemic.
+Previously observed disparities by race/ethnicity still remain.
+Race ratios are predicted to continue to decline within most states.
+The use of Google Trends data to predict 2022 rates is somewhat arbitrary, but is not unreasonable given that many individuals may be searching for information about HIV online after experiencing symptoms or a risk event.
+More information about health seeking behaviors may be necessary to understand the predictions observed in some states.
 
-## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+This report is subject to several limitations.
+The analysis planned to include data from 2008-2022, however 2021-2022 data were not available in time so the analysis was restricted from 2008-2020.
+The prediction model provided reasonable estimates but performed poorly overall.
+Additional state-level data such as census characteristics could be used to improve model fit.
+This analysis also did not evaluate rates for specific transmission groups such as MSM or IDU which may factor heavily into trends for some geographic levels.
 
-## Conclusions
-_What are the main take-home messages?_
+This preliminary analysis summarizes overall trends and highlights several notable disparities in new HIV diagnoses.
+Future analyses should include a wider range of post-COVID data, focus on a small, defined geographic unit, and account for rate differences with and without high-risk transmission groups.
 

### 2023-04-27 Project_Review_Luisi_Wells.qmd
Project Review Wells
addition of project review file from Katie Wells

# Overview

Title of project: "Trends in New HIV Diagnoses in the US, 2008-2020"

Name of project author(s): Nicole Luisi

Name of project reviewer: Katie Wells

# Instructions

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments
The background is very well characterized and adequate context is given as to the purpose of this project thanks to the proper referencing. Clear evidence is given as to why this project was chosen and what it hopes to discover with its exploration and analysis.

### Summary assessment 
* strong contextualization and motivation

## Question description

### Feedback and Comments
Rather than a question being asked, this report evaluates trends in new HIV diagnoses according to race/ethnicity and other subgroups and predicts state-level new diagnoses; it is clear how the data obtained from AidsVu and Google Trends will be used in this evaluation

### Summary assessment
* question/hypotheses fully clear

## Data description

### Feedback and Comments
The data is described very concisely but thoroughly and was documented with the source for further viewing.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis

### Feedback and Comments
The data is very thoroughly cleaned and explored; there is a little documentation for each step though there could be some more. Many figures are produced that explore the trends in the data and gives nice visualization to them. I appreciate that the main figures are in the manuscript and extra are in the exploratory file.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments
While I do not really understand the JoinPoint program, this project made it very easy to understand the analyses done and easy to reproduce with the data without downloading the program. Everything was explained well and concisely.

### Summary assessment
* strong and reasonable analysis

## Presentation

### Feedback and Comments
The results are professional quality and are presented in a way that easily shows the main points of the projects. The colors and size make the graphs easy to interpret. My only recommendation would be to summarize the tables (if possible) to decrease their length. Perhaps the full table could be in the separate file and a shortened one in the manuscript.

### Summary assessment
* results are very well presented

## Discussion/Conclusions

### Feedback and Comments
The findings of the project are clearly discussed and the strengths of the study are concisely presented. They are in line with what I observed when I reproduced the project myself. The limitations discussed are sensible, and the mention of future directions allow for remedy of some of them.

### Summary assessment
* strong, complete and clear discussion

## Further comments

# Overall project content evaluation

## Structure

### Feedback and Comments
The project repository is well organized and easy to follow to reproduce. Everything was labeled sensibly. I found no unnecessary files and was able to follow the readme with ease.

### Summary assessment
* well structured

## Documentation

### Feedback and Comments
There is documentation in each file that is adequate for reproducibilty. I think a few more notes here and there on why each step was chosen would not be amiss, but everything was organized in order so that made it easy to follow each step.

### Summary assessment
* fully and well documented

## Reproducibility

### Feedback and Comments
I was able to reproduce everything by following the readme instructions. I appreciate that you provided a way for reviewers (and others who might try to reproduce your project) to bypass the JoinPoint software while still being able to view your analysis. I had no issues and did not have to make any manual interventions or edits.

### Summary assessment
* fully reproducible without issues

## Thoroughness

### Feedback and Comments
The overall project was very thorough. While we can't see what all happens within the JoinPoint software, the output from it was very thorough and created easy-to-interpret figures. Everything was well-discussed and I was able to see how the outputs of this project related to what was set out to be evaluated.

### Summary assessment
* strong level of thoroughness

## Further comments

This project is the best I have viewed. It is of professional quality and I believe it could be published as-is. It is very thorough and easy to follow, and the reasoning and analysis is very sound. The only thing I would change for improvement is the length of the tables as they take up multiple pages. If they could be abbreviated I think it would make this project perfect.
